Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2025-02-16 demystify.tex
updates

@@

The GRFP website has detailed information on the application process and eligibility requirements. This document focuses on how your application is evaluated and how you can improve your chances of success.

@@

\subsection{Common Pitfalls}\label{sec:grfp-pitfalls}

\paragraph{Too technical} Remember that you're writing to a general audience in CS, not experts in your field.  While NSF tries to match reviewers with your field, CS fields are simply too broad and you might get a reviewer in the field but not in your specific area. 
So avoid technical jargon and explain things in simple terms. If the reviewer does not understand your proposal, they will likely give you a low score.

@@

\paragraph{Preliminary Work} Your research plan should have some preliminary work to convince reviewers that you have thought about the problem and have some initial promising results.  It does not have to a lot, e.g., you don't need published results or an implementation, a small experiment done by hand on some small examples would suffice.  Moreover, it should be concrete and convincing, e.g., \emph{``When being applied to the program in Fig 1. of the paper published PLDI'19 that approach X failed, my idea worked and was able to ...''}.  This will set you apart from others who just have a plan but no results.

\paragraph{Related Work} Many research plans motivate the problem well but do not discuss limitations of existing work, making it questionable if the problem is important or if you are aware of the challenges.  Thus, you should do a thorough literature review and discuss what people have done and their limitations.  Your research plan should then talk about how you plan to address these limitations, i.e., fill the gap in the literature.

\paragraph{Do NOT BS} Many research plans include technical details or proposed work that are vauge, unrealistic, or even impossible.  This is similar to a student taking an exam and write a lot of nonsense to fill the page and hope for partial credits.  As mentioned, reviewers are often chosen based on their expertise in your field and can detect B.S. Again, do your homework and ask someone who knows the field, e.g., your advisor or LoR writer, to review your research plan.

\paragraph{Overexaggeration Lors} Many GRFP applicants are from top schools with well-known professors that are very good at writing. The issue is that they often use flowery language and completely overrate their students. Reviewers are often professors themselves and can sense this overexaggeration.  Moreover, they often know the LoR writers and can compare the LoR with the student's application and see if they match.   So you should ask your LoR writers to be honest, tone down the language (similar to overclaiming research conributions when writing papers), and be specific with concrete examples to demonstrate why you're \emph{``the best''}.

@@

The DoD National Defense Science and Engineering Graduate (NDSEG) is another prestigious fellowship supporting \emph{U.S. citizens} pursuing advanced degrees in STEM fields \emph{critical to national defense} (such as CS).
It provides up to three years of funding, including a  stipend, tuition coverage, and health insurance.

@@

\paragraph{Redacted info.} Unlike the NSF GRFP, for NDSEG reviewers \emph{do not} see your LoRs, i.e., we do not know who wrote them or what they wrote. Moreover, we only see \emph{redacted versions} of your submitted materials (applicants are asked to provide full and redacted versions of your research plan, CV, and other materials). Thus, we do not know what schools your went to, where you did your internships, your publications or research experience.

@@

\begin{commentbox}[Vu:]
I find a bit difficult to evaluate NDSEG applications because I do not have the full picture.  For example, a student mentioned that they have published in the topic they proposed but redacted all information about the publication, so I have no idea where they published, what the paper is about, etc.  
It is also difficult to evaluate the academic background of the student, e.g., some student just started grad school and so their transcript is essentially blank, so I have no idea on their academic background.  It is also annoying to see CVs with so many blanked out information. They might as well should just submit a blank CV one. 
\\

@@

These are some common pitfalls I see in NDSEG applications and some tips to avoid them.  While some of these are similar to NSF GRFP (\autoref{sec:grfp-pitfalls}), some are specific to NDSEG.

@@

    \item Customize your research to DoD, e.g., through specific project from DARPA or ONR. 
    Make this prominent, e.g., in the introduction or in its own section.  This is different from NSF GRFP where you want broader impacts to society.

    \item Unclear research plan. 
     You should aim to have (i) clear problem statement, (ii) why it is important (esp. to DoD), (iii) what has been done and their challenges/limitations, and (iv) what you plan to do to address the challenges.  Moreover, research is often quite technical and ambitious, so you want to show some preliminary work, e.g., a small experiment by hand on some small examples, to convince reviewers that you have thought about the problem and have some initial results.\\

@@

    \item Personal stories and essays should be unique and interesting. Many essays and stories I read are quite generic and do not tell much about the student. For example, being an URM, involving in a robotics club or mentoring underrepresented students are quite common and do not tell much.\\

@@

    \item Not using all allowed pages. Many research plans use only 1/2 or 2/3 of the number of allowed pages, which themselves are already quite short. Such documents are often not detailed enough and do not provide enough information for reviewers to understand and evaluate the research plan.  So use all allowed pages to provide enough details about your research plan. This is especially important as we do not know much about you due to redacted materials.

>>>>

%Securing a prestigious fellowship such as the National Science Foundation Graduate Research Fellowship Program (NSF GRFP) or the Department of Defense National Defense Science and Engineering Graduate Fellowship (DoD NDSEG) can significantly enhance a student’s prospects for admission to a strong PhD program. These fellowships not only provide financial support but also offer academic freedom and prestige that can set an applicant apart from their peers. However, successfully obtaining one of these fellowships is no easy feat—they are highly competitive, and applications are scrutinized rigorously by expert reviewers.

%One of the biggest advantages that applicants from elite institutions often have is access to faculty members who are well-versed in the review process. These faculty members understand the nuances of writing strong letters of recommendation (LoRs) and guiding students through the proposal-writing process. Unfortunately, students from smaller institutions or those with mentors unfamiliar with NSF or DoD fellowship reviews may find themselves at a disadvantage—not because of their potential or research ability, but simply due to a lack of insider knowledge about what makes a strong application.

%This chapter aims to level the playing field by providing an insider’s perspective on how these applications are reviewed. Drawing from my experience as a reviewer for both NSF GRFP and DoD NDSEG, I will demystify the evaluation process, highlight key elements that make applications stand out, and offer strategic advice for applicants and their letter writers. While research experience and proposal quality are, of course, crucial to success, understanding the review criteria and expectations can be just as important.

%Moreover, this chapter will provide guidance for LoR writers, helping them craft letters that effectively support their students’ applications. Many well-intentioned letters fail to address the aspects that reviewers prioritize, which can unintentionally weaken an otherwise strong application. By offering insights into what reviewers look for, I hope to empower both applicants and their mentors to present their strongest possible case for these fellowships.

%Whether you are an aspiring PhD student preparing your application or a faculty member writing a recommendation, this chapter will equip you with the knowledge needed to navigate the NSF GRFP and DoD NDSEG review process with confidence.

@@

The \href{https://www.nsfgrfp.org/}{GRFP website} has detailed information on the application process and eligibility requirements. This document focuses on how your application is evaluated and tips on improving your application.

@@

\subsection{Common Pitfalls and Tips}\label{sec:grfp-pitfalls}

These are common pitfalls that many applicants make in their GRFP applications and tips on how to avoid them.

\item \textbf{Too Technical.} Remember that you're writing to a general audience in CS, not experts in your field.  While NSF tries to match reviewers with your field, CS fields are simply too broad and you might get a reviewer in the field but not in your specific area or familiar with the problem you're working on. 
So avoid technical jargon and explain things in simple terms. If the reviewer cannot understand your proposal, they will likely give you a low score.

@@

\item \textbf{Preliminary Work.} Your research plan should have some preliminary work to convince reviewers that you have thought about the problem and have some initial promising results.  It does not have to a lot, e.g., you don't need published results or an implementation, a small experiment done by hand on some small examples would suffice.  Moreover, it should be concrete and convincing, e.g., \emph{``When being applied to the program in Fig 1. of the paper published PLDI'19 that approach X failed, my idea worked and was able to ...''}.  This will set you apart from others who just have a plan but no results.

\item \textbf{Related Work and Challenges.} Many research plans motivate the problem well but do not discuss limitations of existing work, making it questionable if the problem is important or if you are aware of the challenges.  Thus, you should do a thorough literature review and discuss what people have done and their limitations.  Your research plan should then talk about how you plan to address these limitations, i.e., fill the gap in the literature.

\item \textbf{Do not BS.} Many research plans include technical details or proposed work that are vauge, unrealistic, or even impossible.  This is similar to a student taking an exam and write a lot of nonsense and hope for partial credits.  As mentioned, reviewers are often chosen based on their expertise in your field and can detect B.S. Again, do your homework and ask someone who knows the field, e.g., your advisor or LoR writer, to review your research plan.

\item \textbf{Overexaggeration Lors.} Many GRFP applicants are from top schools with well-known professors that are very good at writing. However, they often use flowery language and completely overrate their students. Reviewers are  professors themselves and can sense this overexaggeration.  So you should ask your LoR writers to tone down the language (similar to overclaiming research conributions when writing papers) and and be specific with concrete examples to demonstrate why you're \emph{``the best''}.

@@

The DoD National Defense Science and Engineering Graduate (NDSEG) is another prestigious fellowship supporting \emph{U.S. citizens} pursuing advanced degrees in STEM fields \emph{critical to national defense} (such as CS).
It provides up to three years of funding, including a  stipend and tuition coverage.

@@

\paragraph{Redacted info.} Unlike the NSF GRFP, reviewers only see \emph{redacted versions} of your submitted materials (applicants are asked to provide full and redacted versions of your research plan, CV, and other materials). Thus, we do not know what schools your went to, where you did your internships, your publications or research experience. We also \emph{do not} see your LoRs, i.e., we do not know who wrote them or what they wrote.

@@

\begin{commentbox}[Vu:]
I find a bit difficult to evaluate NDSEG applications because I do not have the full picture.  For example, a student mentioned that they have published in the topic they proposed but redacted all information about the publication, so I have no idea where they published, what the paper is about, etc.  
It is also difficult to evaluate the academic background of the student, e.g., some student just started grad school and so their transcript is essentially blank. It is also annoying to see CVs with so many blanked out information. They might as well should just submit a blank CV one. 
\\

@@

These are some common pitfalls I see in NDSEG applications and some tips to avoid them.  Some of these are similar to NSF GRFP (\autoref{sec:grfp-pitfalls}) while some are specific to NDSEG.

@@

    \item \textbf{Customize your research to DoD}, e.g., through specific project from DARPA or ONR. 
    Make this prominent, e.g., in the introduction or in its own section.  This is different from NSF GRFP where you want broader impacts to society.

    \item \textbf{Unclear research plan.}
     You should aim to have (i) clear problem statement, (ii) why it is important (esp. to DoD), (iii) what has been done and their challenges/limitations, and (iv) what you plan to do to address the challenges.  Moreover, research is often quite technical and ambitious, so you want to show some preliminary work, e.g., a small experiment by hand on some small examples, to convince reviewers that you have thought about the problem and have some initial results.

@@

    \item \textbf{Personal stories and essays should be unique and interesting}. Many essays and stories I read are quite generic and do not tell much about the student. For example, being an URM, involving in a robotics club or mentoring underrepresented students are quite common and do not tell much.

@@

    \item \textbf{Not using all allowed pages}. Many research plans use only 1/2 or 2/3 of the number of allowed pages, which themselves are already quite short. Such documents are often not detailed enough and do not provide enough information for reviewers to understand and evaluate the research plan.  So use all allowed pages to provide enough details about your research plan. This is especially important as we do not know much about you due to redacted materials.
